Towards Practical First-Order Model Counting

### 2025-06-05 doc/paper/paper.tex
Cleanup
Potentially the final (camera-ready) version

\documentclass[a4paper,UKenglish,cleveref,autoref,thm-restate]{lipics-v2021}
\pdfoutput=1
\bibliographystyle{plainurl}

@@

\author{Ananth K. Kidambi}{Indian Institute of Technology Bombay, Mumbai, India}{210051002@iitb.ac.in}{}{}
\author{Guramrit Singh}{Indian Institute of Technology Bombay, Mumbai, India}{guramrit@iitb.ac.in}{}{}
\author{Paulius Dilkas}{University of Toronto, Toronto, Canada \and Vector Institute, Toronto, Canada \and \url{https://dilkas.github.io/}}{paulius.dilkas@utoronto.ca}{https://orcid.org/0000-0001-9185-7840}{}
\author{Kuldeep S. Meel}{University of Toronto, Toronto, Canada \and \url{https://www.cs.toronto.edu/~meel/}}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{}

@@

@@

\acknowledgements{The first two authors contributed equally. Part of the
  research was conducted while all authors were at the National University of
  Singapore.}
\nolinenumbers

@@

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{siunitx}
\usepackage{tikz}
\usepackage{microtype}
\usepackage{mathtools}

@@

\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Cranetwo}{\textsc{Gantry}}
\newcommand{\Cranebfs}{\textsc{Gantry-BFS}}
\newcommand{\Cranegreedy}{\textsc{Gantry-Greedy}}
\newcommand{\friends}{\emph{Friends \& Smokers}}
\newcommand{\functions}{\emph{Functions}}
\newcommand{\bijections}{\emph{Bijections}}

@@

\sisetup{group-separator = {,}}
\newtheorem{assumption}[theorem]{Assumption}

@@

FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to cell
enumeration~\cite{DBLP:conf/uai/BremenK21}, local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, and Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}. (See the technical appendix for a
more detailed comparison of the state-of-the-art FOMC algorithms.) FOKC-based
algorithms are particularly prominent, transforming sentences into structured
representations such as circuits or graphs. Even when multiple algorithms can
solve the same instance, FOKC algorithms are known to find polynomial-time
solutions where the polynomial has a lower degree than other
approaches~\cite{DBLP:conf/kr/DilkasB23}. The recently developed ability of a
FOKC algorithm to formulate solutions in terms of recursive
functions~\cite{DBLP:conf/kr/DilkasB23} is also noteworthy since the only other
proposed alternative is to guess recursive
relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}. Notable examples of FOKC
algorithms include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its
extension \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}.

@@

We begin this section by describing some notation that we will use throughout
the paper. Then, in \cref{sec:logic,sec:fomc}, we introduce the basic
terminology of first-order logic and formally define (W)FOMC\@. \Cref{sec:crane}
outlines the principles of FOKC, particularly in the context of \textsc{Crane}.
Finally, in \cref{sec:algebra}, we introduce the algebraic terminology used to
describe the output of \textsc{Crane}, i.e., functions and equations that define
them.

\paragraph*{Notation}
We use $\mathbb{N}_{0}$ to represent the set of non-negative integers. In both
algebra and logic, we write $S\sigma$ to denote the application of a
\emph{substitution} $\sigma$ to an expression $S$, where
$\sigma = [x_{1} \mapsto y_{1}, x_{2} \mapsto y_{2}, \dots, x_{n} \mapsto y_{n}]$
signifies the replacement of all instances of $x_{i}$ with $y_{i}$ for all
$i = 1, \dots, n$. Additionally, for any variable $n$ and
$a, b \in \mathbb{N}_{0}$, let $[a \le n \le b] \coloneqq \begin{cases}
  1 & \text{if $a \le n \le b$} \\
  0 & \text{otherwise}
\end{cases}$.

@@

\begin{definition}[Structure, model]\label{def:model}
  Let $\phi$ be a sentence. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains. Let
  $\sigma$ be a map from the domains of $\phi$ to their interpretations as
  finite sets, ensuring the sets are pairwise disjoint and contain the
  corresponding constants from $\phi$. A \emph{structure} of $\phi$ is a set $M$
  of ground literals defined by adding to $M$ either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and every $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it makes $\phi$ valid.
\end{definition}

@@

\begin{remark*}
  The distinctness of domains is significant in two respects. First, in terms of
  expressiveness, a clause such as $\forall x \in \Delta\text{. }P(x, x)$ is
  valid if predicate $P$ operates over two copies of the same domain and invalid
  otherwise. Second, having more distinct domains makes the problem more
  decomposable for the FOKC algorithm. With distinct domains, the algorithm can
  make assumptions or deductions about, e.g., the first domain of predicate $P$
  without worrying how (or if) they apply to the second domain.
\end{remark*}

@@

@@

This stage transforms an arbitrary sentence into the format described in
\cref{sec:logic}, primarily by eliminating existential quantifiers. For example,
the first conjunct of \cref{eq:bijections}, i.e.,
\begin{equation}\label[formula]{eq:skolemizationinitial}
  \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
\end{equation}
is transformed into
\begin{equation}\label[formula]{eq:skolemization}
  \begin{aligned}
    (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
    (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
    (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
    (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)),
  \end{aligned}
\end{equation}
where $Z/1$ and $S/1$ are two new predicates with $w^{-}(S) = -1$. One can
verify that the WFOMC of \cref{eq:skolemizationinitial,eq:skolemization} is the
same.

@@

After preprocessing, \textsc{Crane} compiles the sentence into the triple
$(\mathcal{E}, \mathcal{F}, \mathcal{D})$, where $\mathcal{E}$ is the set of
equations, and $\mathcal{F}$ and $\mathcal{D}$ are auxiliary maps. $\mathcal{F}$
maps function names to sentences. $\mathcal{D}$ maps function names and argument
indices to domains. $\mathcal{E}$ can contain any number of functions, one of
which (which we will always denote as $f$) represents the solution to the FOMC
problem. Computing the FOMC for specific domain sizes involves invoking $f$ with
those sizes as inputs. $\mathcal{D}$ records this correspondence between
function arguments and domains.

@@

\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node[anchor=west] at (-1, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (compilewithbasecases) {\CompileWithBaseCases};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compile to C++};

@@

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{Using \Cranetwo{} to compute the model count of a sentence $\phi$.
    First, \Cranetwo{} compiles $\phi$ into a set of equations, which form the
    basis for creating a C++ program. Executing the program with different
    command line arguments calculates the model count of $\phi$ for various
    domain sizes. To accomplish this, the \CompileWithBaseCases procedure
    employs \textsc{Crane}, algebraic simplification techniques (denoted as
    \Simplify), and two other auxiliary procedures. }\label{fig:overview}
\end{figure*}

@@

\cref{sec:completing} also describes the \Simplify procedure for algebraic
simplification. It is crucial for simplifying, e.g., a sum of $n$ terms, only
some of which are non-zero. More generally, the equations returned by
\textsc{Crane} often benefit from easily detectable algebraic simplifications
such as $0 \cdot \text{anything} = 0$ and $\text{anything}^{0} = 1$.

@@

In the rest of this section, we will show that the base cases identified by
\FindBaseCases are sufficient for the algorithm to terminate.\footnote{Note that
  characterising the fine-grained complexity of the solutions found by
  \Cranetwo{} or other FOMC algorithms is an emerging area of research. These
  questions have been partially addressed in previous
  work~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/kr/TothK24} and are unrelated to
  the goals of this section.} Let $\mathcal{E}$ denote the equations returned by
\CompileWithBaseCases.

\begin{theorem}\label{thm:halting}
  Given an $n$-ary function $f$ in $\mathcal{E}$ and
  $\mathbf{x} \in \mathbb{N}_{0}^{n}$, the evaluation of $f(\mathbf{x})$
  terminates.
\end{theorem}

We prove \cref{thm:halting} using double induction. First, we apply induction to
the number of functions in $\mathcal{E}$. Then, we use induction on the arity of
the `last' function in $\mathcal{E}$ according to a topological ordering. Before
proving \cref{thm:halting}, we make a few observations about this and
previous~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/ijcai/BroeckTMDR11} work.

@@

\Cref{assumption2} prevents mutual recursion and other cyclic scenarios.

@@

@@

@@

In this section, we will describe the final step of \Cranetwo{} as outlined in
\cref{fig:overview}, i.e., translating the set of equations $\mathcal{E}$ into
C++ code. Recall that this step is crucial for the usability of the algorithm;
otherwise, function definitions would remain purely mathematical, with no
convenient way to compute the model count for particular domain sizes. Once a
C++ program is produced, it can be executed with different command-line
arguments to determine the model count of the sentence for various domain sizes.

@@

The experiments used an Intel Skylake \SI{2.4}{\giga\hertz} CPU with
\SI{188}{\gibi\byte} of memory and CentOS~7. We used the Intel C++ Compiler
2020u4 for C++ programs, Julia~1.10.4 for \textsc{FastWFOMC}, and the Java
Virtual Machine 1.8.0\_201 for \textsc{ForcLift} and \Cranetwo{}. Although
implemented in different languages, \Cranetwo{} and \textsc{FastWFOMC} use the
same GNU Multiple Precision Arithmetic Library for arbitrary-precision
arithmetic.

We ran each algorithm on each benchmark using domains of size
$2^{1}, 2^{2}, 2^{3}$, and so on until an algorithm failed to handle a domain
size due to a timeout (of one hour) or out-of-memory or out-of-precision errors.
While we separately measured compilation and inference time, we primarily focus
on total runtime, dominated by the latter. We verified the accuracy of the
numerical answers using the corresponding integer sequences in the On-Line
Encyclopedia of Integer Sequences~\cite{oeis}.

@@

The three benchmarks cover a wide range of possibilities. The \friends{}
benchmark uses multiple predicates and only two variables without employing any
constructs specific to only one of the (W)FOMC algorithms. The \functions{}
benchmark, on the other hand, can still be handled by all the algorithms, but
its formulation relies on algorithm-specific constructs. Namely, for
\textsc{ForcLift} and \Cranetwo{}, the sentence is written using three
variables. Since \textsc{FastWFOMC} supports at most two variables, we rewrite
the \cref{eq:functions} with two variables and a \emph{cardinality constraint}
(see the technical appendix for details about translating the benchmarks to
other first-order logics). Lastly, the \bijections{} benchmark is an example of
a sentence that \textsc{FastWFOMC} can handle but \textsc{ForcLift} cannot.

@@

\begin{figure*}[t]
  \centering
  \includegraphics{plot.pdf}
  \caption{The runtime of the algorithms as a function of domain size. Note that
    both axes are on a logarithmic scale.}\label{fig:plot}
\end{figure*}

@@

One might notice that the runtime of \textsc{FastWFOMC} and \textsc{ForcLift} is
slightly higher for the smallest domain size. This peculiarity is the
consequence of \emph{just-in-time} (JIT) compilation. As \Cranetwo{} is only run
once per benchmark, we include the JIT compilation time in its overall runtime
across all domain sizes. Additionally, while the compilation time of
\textsc{ForcLift} is generally lower compared to \Cranetwo{}, neither
significantly affects the overall runtime. Specifically, \textsc{ForcLift}
compilation typically takes around \SI{0.5}{\second}, while \Cranetwo{}
compilation takes around \SI{2.3}{\second}.

>>>>

\documentclass[a4paper,UKenglish,cleveref,autoref]{lipics-v2021}
\pdfoutput=1
\bibliographystyle{plainurl}

@@

\author{Ananth K. Kidambi}{Indian Institute of Technology Bombay, Mumbai, India}{210051002@iitb.ac.in}{}{}
\author{Guramrit Singh}{Indian Institute of Technology Bombay, Mumbai, India}{guramrit@iitb.ac.in}{}{}
\author{Paulius Dilkas}{University of Toronto, Toronto, Canada \and Vector Institute, Toronto, Canada \and \url{https://dilkas.github.io/}}{paulius.dilkas@utoronto.ca}{https://orcid.org/0000-0001-9185-7840}{}
\author{Kuldeep S. Meel}{Georgia Institute of Technology, Atlanta, GA, USA \and University of Toronto, Toronto, Canada \and \url{https://www.cs.toronto.edu/~meel/}}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{}

@@

\supplementdetails[subcategory={Source Code}]{Software}{https://doi.org/10.5281/zenodo.15583934}

@@

\acknowledgements{The first two authors contributed equally. Part of the
  research was conducted while all authors were at the National University of
  Singapore. We express our gratitude to the anonymous reviewers for their
  valuable comments and suggestions.}

\nolinenumbers

@@

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{siunitx}
\usepackage{tikz}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage[table]{xcolor}
\usepackage{pifont}

@@

\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Cranetwo}{\textsc{Crane2}}
\newcommand{\Cranebfs}{\textsc{Crane2-BFS}}
\newcommand{\Cranegreedy}{\textsc{Crane2-Greedy}}
\newcommand{\friends}{\emph{Friends \& Smokers}}
\newcommand{\functions}{\emph{Functions}}
\newcommand{\bijections}{\emph{Bijections}}

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{CC}$}

@@

\sisetup{group-separator = {,}}
\sisetup{range-phrase=--}

\newtheorem{assumption}[theorem]{Assumption}

@@

FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to cell
enumeration~\cite{DBLP:conf/uai/BremenK21}, local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, and Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}. FOKC-based algorithms are
particularly prominent, transforming sentences into structured representations
such as circuits or graphs. Even when multiple algorithms can solve the same
instance, FOKC algorithms are known to find polynomial-time solutions where the
polynomial has a lower degree than other
approaches~\cite{DBLP:conf/kr/DilkasB23}. The recently developed ability of a
FOKC algorithm to formulate solutions in terms of recursive
functions~\cite{DBLP:conf/kr/DilkasB23} is also noteworthy since the only other
proposed alternative is to guess recursive
relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}. Notable examples of FOKC
algorithms include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its
extension \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}.

@@

We begin this section by describing some notation that we will use throughout
the paper. Then, in \cref{sec:logic,sec:fomc}, we introduce the basic
terminology of first-order logic and formally define (W)FOMC\@.
\Cref{sec:threelogics} provides a brief overview of the different types of
first-order logics commonly used in FOMC literature. \Cref{sec:crane} outlines
the principles of FOKC, particularly in the context of \textsc{Crane}. Finally,
in \cref{sec:algebra}, we introduce the algebraic terminology used to describe
the output of \textsc{Crane}, i.e., functions and equations that define them.

\paragraph*{Notation}
We use $\mathbb{N}_{0}$ to represent the set of non-negative integers. In both
algebra and logic, we write
$S[x_{1} \mapsto y_{1}, x_{2} \mapsto y_{2}, \dots, x_{n} \mapsto y_{n}]$ to
denote the application of a \emph{substitution} to an expression $S$, where all
instances of $x_{i}$ are replaced with $y_{i}$ for all $i = 1, \dots, n$.
Additionally, for any variable $n$ and $a, b \in \mathbb{N}_{0}$, let
$[a \le n \le b] \coloneqq \begin{cases}
  1 & \text{if $a \le n \le b$} \\
  0 & \text{otherwise}
\end{cases}$.

@@

\begin{definition}[Structure, model]\label{def:model}
  Let $\phi$ be a sentence. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains. Let
  $\sigma$ be a map from the domains of $\phi$ to their interpretations as
  finite sets, ensuring the sets are pairwise disjoint and contain the
  corresponding constants from $\phi$. A \emph{structure} of $\phi$ is a set $M$
  of ground literals. It is defined by adding to $M$ either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and every $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. (Here, $\prod$ is the
  usual Cartesian product.) A structure is a \emph{model} if it makes $\phi$
  valid.
\end{definition}

@@

@@

\subsection{The Three Logics of FOMC}\label{sec:threelogics}

\begin{table}[t]
  \centering
  \caption{A comparison of the three logics used in FOMC\@. The
    2\textsuperscript{nd}--5\textsuperscript{th} columns refer to the number of
    sorts, support for constants, the maximum number of variables, and supported
    quantifiers, respectively. The last column lists supported atoms in addition
    to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and an $n$-tuple
    of terms $\mathbf{t}$. Here, $k$ and $m$ are non-negative integers, where
    $m$ depends on the domain size, $P$ is a predicate, and $x$ and $y$ are
    terms.}\label{tbl:logics}
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Add.\ atoms\\
    \midrule
    \rowcolor{gray!10}
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & \multirow{2}{*}{one} & \multirow{2}{*}{\xmark} & \multirow{2}{*}{two} & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & & & & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
\end{table}

FOMC commonly utilises three types of first-order logic: \FO{}, \Ctwo{}, and
\UFO{}. \Cref{tbl:logics} summarises the key differences among them. \FO{} is
the input format for \textsc{ForcLift} and its extensions \textsc{Crane} and
\Cranetwo{}. \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
related methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}.
(Note that no algorithm accepts \Ctwo{} as input.) Finally, \UFO{} is the input
format supported by the most recent implementation of
\textsc{FastWFOMC}~\cite{DBLP:conf/kr/TothK24}. All three logics are
function-free, and domains are always assumed to be finite. As usual, we
presuppose the \emph{unique name assumption}, which states that two constants
are equal if and only if they are the same constant~\cite{DBLP:books/aw/RN2020}.

In \FO{}, each term has a designated \emph{sort}, and each predicate $P/n$
corresponds to a sequence of $n$ sorts. Each sort has its corresponding domain.
These assignments to sorts are typically left implicit and follow from the
quantifiers, e.g., $\forall x,y \in \Delta\text{. }P(x, y)$ implies that the
variables $x$ and $y$ have the same sort. On the other hand,
$\forall x \in \Delta\text{. }\forall y \in \Gamma\text{. } P(x, y)$ implies
that $x$ and $y$ have different sorts, and it would be improper to write, for
example, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{.
} P(x, y) \lor x = y$. \FO{} is also the only logic to support constants,
sentences with more than two variables, and the equality predicate. While we do
not explicitly refer to sorts in the rest of the paper, the many-sorted nature
of \FO{} is paramount to the algorithms presented therein.

\begin{remark*}
  In the case of \textsc{ForcLift} and its extensions, support for a sentence as
  valid input does not imply that the algorithm can compile the sentence into a
  circuit or graph suitable for lifted model counting. However,
  \textsc{ForcLift} compilation always succeeds on any \FO{} sentence without
  constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}.
\end{remark*}

Compared to \FO{}, \Ctwo{} and \UFO{} lack support for constants, the equality
predicate, multiple domains, and sentences with more than two variables. The
advantage that \Ctwo{} brings over \FO{} is the inclusion of \emph{counting
  quantifiers}. That is, alongside $\forall$ and $\exists$, \Ctwo{} supports
$\exists^{=k}$, $\exists^{\le k}$, and $\exists^{\ge k}$ for any positive
integer $k$. For example, $\exists^{=1} x\text{. }\phi(x)$ means that there
exists \emph{exactly one} $x$ such that $\phi(x)$, and $\exists^{\le 2} x\text{.
}\phi(x)$ means that there exist \emph{at most two} such $x$. \UFO{}, on the
other hand, does not support any existential quantifiers but instead
incorporates \emph{(equality) cardinality constraints}. For example, $|P| = 3$
constrains all models to have \emph{precisely three positive literals with the
  predicate $P$}. Except when stated otherwise, we will write all example
sentences in \FO{}.

@@

This stage transforms an arbitrary sentence into the format described in
\cref{sec:logic}, primarily by eliminating existential quantifiers via a process
called \emph{Skolemization}~\cite{DBLP:conf/kr/BroeckMD14}. For example, the
first conjunct of \cref{eq:bijections}, i.e.,
\begin{equation}\label[formula]{eq:skolemizationinitial}
  \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
\end{equation}
is transformed into
\begin{equation}\label[formula]{eq:skolemization}
  \begin{aligned}
    (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
    (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
    (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
    (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)),
  \end{aligned}
\end{equation}
where $Z/1$ and $S/1$ are two new predicates with $w^{-}(S) = -1$. The idea
behind Skolemization is that the output sentence has more models, but for every
additional model with a positive weight, there is an equivalent model with a
negative weight. Hence, the WFOMC of
\cref{eq:skolemizationinitial,eq:skolemization} is the same. Moreover, note that
in this example, preprocessing introduces some redundancy, i.e., the second and
third clauses of \cref{eq:skolemization} could be removed. These redundancies
are handled at the beginning of the compilation process described in
\cref{sec:compilation}.

To illustrate how Skolemization preserves the WFOMC, we let $\Gamma = \{\,a\,\}$
and $\Delta = \{\,1, 2\,\}$ and assume that $w^+(P) = w^-(P) = 1$. Here, for
conciseness, we omit negative literals when writing models.
\Cref{eq:skolemizationinitial} has three models: $\{\,P(a, 1)\,\}$,
$\{\,P(a, 2)\,\}$, and $\{\,P(a, 1), P(a, 2)\,\}$, all with weight 1, so the
WFOMC is 3. \Cref{eq:skolemization} has the following models:
\begin{itemize}
  \item $\{\,Z(a)\,\}$,
  \item $\{\,Z(a), S(a)\,\}$,
  \item $\{\,Z(a), S(a), P(a, 1)\,\}$,
  \item $\{\,Z(a), S(a), P(a, 2)\,\}$,
  \item $\{\,Z(a), S(a), P(a, 1), P(a, 2)\,\}$.
\end{itemize}
The first model has a weight of -1 because $w^-(S) = -1$, while all the other
models have a weight of 1. Hence, the WFOMC is $4 - 1 = 3$.

@@

After preprocessing, \textsc{Crane} compiles the sentence into the triple
$(\mathcal{E}, \mathcal{F}, \mathcal{D})$, where $\mathcal{E}$ is the set of
equations, and $\mathcal{F}$ and $\mathcal{D}$ are auxiliary maps. $\mathcal{F}$
maps function names to sentences in such a way that the evaluation of the former
corresponds to the FOMC of the latter. $\mathcal{D}$ maps function names and
argument indices to domains. The equations in $\mathcal{E}$ can define an
arbitrary number of functions, one of which (which we will always denote as $f$)
represents the solution to the FOMC problem. Computing the FOMC for specific
domain sizes involves invoking $f$ with those sizes as inputs. $\mathcal{D}$
records this correspondence between function arguments and domains.

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[anchor=west] at (-1, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (compilewithbasecases) {\CompileWithBaseCases};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compile to C++};

@@

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{Using \Cranetwo{} to compute the model count of a sentence $\phi$.
    First, \Cranetwo{} compiles $\phi$ into a set of equations, which form the
    basis for creating a C++ program. Executing the program with different
    command line arguments calculates the model count of $\phi$ for various
    domain sizes. To accomplish this, the \CompileWithBaseCases procedure
    employs \textsc{Crane}, algebraic simplification techniques (denoted as
    \Simplify), and two other auxiliary procedures. }\label{fig:overview}
\end{figure}

@@

\cref{sec:completing} also describes the \Simplify procedure for algebraic
simplification. It is crucial for simplifying, e.g., a sum of $n$ terms, only
some of which are non-zero. More generally, the equations returned by
\textsc{Crane} often benefit from easily detectable algebraic simplifications
such as $0 \cdot \text{anything} = 0$ and $\text{anything}^{0} = 1$. Note that
the latter includes $0^{0} = 1$, as 1 is the multiplicative identity.

@@

In the rest of this section, we will show that the base cases identified by
\FindBaseCases are sufficient for the algorithm to terminate.\footnote{Note that
  characterising the fine-grained complexity of the solutions found by
  \Cranetwo{} or other FOMC algorithms is an emerging area of research. These
  questions have been partially addressed in previous
  work~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/kr/TothK24} and are unrelated to
  the goals of this section.}

\begin{theorem}\label{thm:halting}
  Let $\mathcal{E}$ denote the set of equations returned by
  \CompileWithBaseCases, and let $f$ be one of the functions defined in
  $\mathcal{E}$. The evaluation of $f(\mathbf{x})$ terminates for any tuple
  $\mathbf{x}$ of non-negative integers (of appropriate length).
\end{theorem}

We prove \cref{thm:halting} using double induction. First, we apply induction to
the number of functions in $\mathcal{E}$. Then, we use induction on the arity of
the `last' function in $\mathcal{E}$ according to a topological ordering. Before
proving \cref{thm:halting}, we make a few observations that are immediate
consequences of this and previous
work~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/ijcai/BroeckTMDR11}.

@@

\Cref{assumption2} prevents mutual recursion and other cyclic scenarios. It
arises from the timeline of FOKC and the fact that, while constructing a graph,
the algorithm cannot draw an edge to a vertex that does not yet exist.

@@

Indeed, each $\expr$ is defined by a node type, all of which are finite
computations listed in previous
work~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}.

@@

Now that we have handled the special case when $\mathcal{E}$ defines one
function, we are ready to prove the main theorem.

@@

In this section, we will describe the final step of \Cranetwo{} as outlined in
\cref{fig:overview}, i.e., translating the set of equations $\mathcal{E}$ into
C++ code. (Note that the compilation does not rely on any unique features of
C++; the equations could easily be compiled into any other programming
language.) Recall that this step is crucial for the usability of the algorithm;
otherwise, function definitions would remain purely mathematical, with no
convenient way to compute the model count for particular domain sizes. Once a
C++ program is produced, it can be executed with different command-line
arguments to determine the model count of the sentence for various domain sizes.
A typical C++ program for one of the benchmarks in \cref{sec:experiments}
contains about \numrange{80}{100} lines of code. Moreover, although we focus on
FOMC in this work, support for WFOMC could easily be added by using
floating-point numbers instead of integers.

@@

The experiments used an Intel Skylake \SI{2.4}{\giga\hertz} CPU with
\SI{188}{\gibi\byte} of memory and CentOS~7. We used the Intel C++ Compiler
2020u4 for C++ programs, Julia~1.10.4 for \textsc{FastWFOMC}, and the Java
Virtual Machine 1.8.0\_201 for \textsc{ForcLift} and \Cranetwo{}. Although
implemented in different languages, \Cranetwo{} and \textsc{FastWFOMC} use the
same GNU Multiple Precision Arithmetic Library for arbitrary-precision
arithmetic. Additionally, while \textsc{ForcLift} could be extended to support
arbitrary-precision arithmetic, its compilation rules are not extensive enough
to handle some of the benchmarks.

We ran each algorithm on each benchmark using domains of size
$2^{1}, 2^{2}, 2^{3}$, and so on until an algorithm failed to handle a domain
size due to a timeout (of one hour) or out-of-memory or out-of-precision errors.
We verified the accuracy of the numerical answers using the corresponding
integer sequences in the On-Line Encyclopedia of Integer Sequences~\cite{oeis}.

@@

\paragraph*{Our Benchmarks in \Ctwo{} and \UFO{}}
For completeness and reproducibility, let us translate the benchmark sentences
from \FO{} to \Ctwo{} and \UFO{}. Since \friends{} is a relatively simple
sentence, it remains the same in \Ctwo{} and \UFO{}. For \functions{}, in
\Ctwo{}, one would write
\[
  \forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{. }P(x, y).
\]
In \UFO{}, the equivalent formulation is
\begin{equation}\label[formula]{eq:functions1}
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)) \land (|P| = |\Delta|),
\end{equation}
where $w^{-}(S) = -1$. Although \cref{eq:functions1} has more models than its
counterpart in \Ctwo{}, the negative weight $w^{-}(S) = -1$ causes some of the
terms in the definition of WFOMC to cancel out. The translation of \bijections{}
is similar to that of \functions{}. In \Ctwo{}, one could write
\[
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{.
  }P(x, y)) \land (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta\text{.
  }P(x, y)).
\]
Similarly, in \UFO{}, the equivalent formulation is
\[
  (\forall x, y \in \Delta\text{.
  }R(x) \lor \neg P(x, y)) \land (\forall x, y \in \Delta\text{.
  }S(x) \lor \neg P(y, x)) \land (|P| = |\Delta|),
\]
where $w^{-}(R) = w^{-}(S) = -1$.

The three benchmarks cover a wide range of possibilities. The \friends{}
benchmark uses multiple predicates and only two variables without employing any
constructs specific to only one of the (W)FOMC algorithms. The \functions{}
benchmark, on the other hand, can still be handled by all the algorithms, but
its formulation relies on algorithm-specific constructs. Namely, for
\textsc{ForcLift} and \Cranetwo{}, the sentence is written using three
variables. Since \textsc{FastWFOMC} supports at most two variables, we rewrite
the \cref{eq:functions} with two variables and a \emph{cardinality constraint}.
Lastly, the \bijections{} benchmark is an example of a sentence that
\textsc{FastWFOMC} can handle but \textsc{ForcLift} cannot.

@@

\begin{figure}[t]
  \centering
  \includegraphics{plot.pdf}
  \caption{The (total) runtime of the algorithms as a function of domain size.
    Note that both axes are on a logarithmic scale.}\label{fig:plot}
\end{figure}

@@

\begin{table}[t]
  \caption{The compilation times of FOKC algorithms, sorted in ascending
    order.}\label{tbl:compilation}
  \centering
  \begin{tabular}{llr}
    \toprule
    Algorithm & Benchmark & Compilation time (s)\\
    \midrule
    \rowcolor{gray!10}
              & \functions{} & 0.44\\
    \rowcolor{gray!10}
    \multirow{-2}{*}{\textsc{ForcLift}} & \friends{} & 0.52\\
              & \friends{} & 2.20\\
    \multirow{-2}{*}{\Cranegreedy{}} & \functions{} & 2.36\\
    \rowcolor{gray!10}
              & \friends{} & 2.23\\
    \rowcolor{gray!10}
              & \functions{} & 2.53\\
    \rowcolor{gray!10}
    \multirow{-3}{*}{\Cranebfs{}} & \bijections{} & 7.70\\
  \end{tabular}
\end{table}

One might notice that the runtime of \textsc{FastWFOMC} and \textsc{ForcLift} is
slightly higher for the smallest domain size. This peculiarity is the
consequence of \emph{just-in-time} (JIT) compilation. As \Cranetwo{} is only run
once per benchmark, we include the JIT compilation time in its overall runtime
across all domain sizes. Additionally, \cref{tbl:compilation} summarises the
compilation time of each FOKC algorithm on each suitable benchmark. While the
compilation time of \textsc{ForcLift} is generally lower compared to
\Cranetwo{}, neither significantly affects the overall runtime.
